---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  tidy = "styler"
)
```

# SpatialGEV

<!-- badges: start -->
[![](https://www.r-pkg.org/badges/version/SpatialGEV)](https://cran.r-project.org/package=SpatialGEV)
[![R-CMD-check](https://github.com/meixichen/SpatialGEV/workflows/R-CMD-check/badge.svg)](https://github.com/meixichen/SpatialGEV/actions)
<!-- badges: end -->

*Meixi Chen, Martin Lysy, Reza Ramezan*

---

## Description

A fast Bayesian inference method for spatial random effects modelling of weather extremes. 
The latent spatial variables are efficiently marginalized by a Laplace approximation using the 
[***TMB***](https://github.com/kaskr/adcomp) library, which leverages efficient automatic 
differentiation in C++. The models are compiled in C++, whereas the optimization step is carried 
out in R. With this package, users can fit spatial GEV models with different complexities to 
their dataset without having to formulate the model using C++.  This package also offers a 
method to sample from the approximate posterior distributions of both fixed and random effects, 
which will be useful for downstream analysis.

## Installation

Before installing ***SpatialGEV***, make sure you have ***TMB*** installed following the instructions [here](https://github.com/kaskr/adcomp/wiki/Download). 

***SpatialGEV*** uses several functions from the ***INLA*** package for SPDE approximation to the Matérn covariance as well as mesh creation on the spatial domain. If the user would like to use the SPDE method (i.e. `kernel="spde"` in `spatialGEV_fit()`), please first install package ***INLA***. Since ***INLA*** is not on CRAN, it needs to be downloaded following their instruction [here](https://www.r-inla.org/download-install).

To download the stable version of this package, run
```{r install-pkg-cran, eval=FALSE}
install.packages("SpatialGEV")
```

To download the development version of this package, run 
```{r install-pkg-github, eval=FALSE}
devtools::install_github("meixichen/SpatialGEV")
```

## Example

Using the simulated data set `simulatedData2` provided in the package, we demonstrate how to use this package. Spatial variation of the GEV parameters are plotted below.

```{r show-data, fig.height=4, fig.width=4.5, out.width="50%"}
library(SpatialGEV)
# GEV parameters simulated from Gaussian random fields
a <- simulatedData2$a # location
logb <- simulatedData2$logb # log scale
logs <- simulatedData2$logs # log shape
locs <- simulatedData2$locs # coordinate matrix
n_loc <- nrow(locs) # number of locations
y <- Map(evd::rgev, n=sample(50:70, n_loc, replace=TRUE),
         loc=a, scale=exp(logb), shape=exp(logs)) # observations

filled.contour(
  x = unique(locs$x),
  y = unique(locs$y),
  z = matrix(a, ncol=sqrt(n_loc)), 
  color.palette = terrain.colors,
  xlab="Longitude", ylab="Latitude", 
  main="Spatial variation of a",
  cex.lab=1,cex.axis=1
)

filled.contour(
  x = unique(locs$x),
  y = unique(locs$y),
  z = matrix(exp(logb), ncol=sqrt(n_loc)), 
  color.palette = terrain.colors,
  xlab="Longitude", ylab="Latitude", 
  main="Spatial variation of b",
  cex.lab=1,cex.axis=1
)

filled.contour(
  x = unique(locs$x),
  y = unique(locs$y),
  z = matrix(exp(logs), ncol=sqrt(n_loc)),
  color.palette = terrain.colors,
  xlab="Longitude", ylab="Latitude",
  main="Spatial variation of s",
  cex.lab=1,cex.axis=1
)
```

To fit a GEV-GP model to the simulated data, use the `spatialGEV_fit()` function. We use `random="abs"` to indicate that all three GEV parameters are treated as random effects. The shape parameter `s` is constrained to be positive (log transformed) by specifying `reparam_s="positive"`. The covariance kernel function used here is the SPDE-approximated Matérn kernel `kernel="spde"`. Initial parameter values are passed to `init_param` using a list.

```{r fit}
fit <- spatialGEV_fit(
  data = y, locs = locs, random = "abs",
  init_param = list(
    a = rep(60, n_loc),
    log_b = rep(2,n_loc),
    s = rep(-3,n_loc),
    beta_a = 60, beta_b = 2, beta_s = -2,
    log_sigma_a = 1.5, log_kappa_a = -2,
    log_sigma_b = 1.5, log_kappa_b = -2,
    log_sigma_s = -1, log_kappa_s = -2
  ),
  reparam_s = "positive",
  kernel="spde",
  silent = TRUE
)

class(fit)
print(fit)
```

Posterior samples of the random and fixed effects are drawn using `spatialGEV_sample()`. Specify `observation=TRUE` if we would also like to draw from the posterior predictive distribution.

```{r sample}
sam <- spatialGEV_sample(model = fit, n_draw = 1e4, observation = TRUE)
print(sam)
```

To get summary statistics of the posterior samples, use `summary()` on the sample object.
```{r sample-summary}
pos_summary <- summary(sam)
pos_summary$param_summary[1:5,]
pos_summary$y_summary[1:5,]
```

One can also plot the full posteriors using e.g., the [***bayespolot***](https://mc-stan.org/bayesplot/) package.
```{r sample-plot}
library(bayesplot)
library(ggplot2)
mcmc_areas(
  x = sam$parameter_draws[,1:5],
  prob = .95,
  point_est = "mean"
) +
  ggtitle(
    "Posterior distributions of a1 - a5",
    "with posterior means and 95% credible intervals"
  )
```

# TODO

- [ ] Consider a shorter name, e.g., `sgev_*`, than `spatialGEV_*`.

- [ ] Argument `init_params` adds a lot of complexity to `spatialGEV_fit()`.  Perhaps this function can be broken down into two parts: `spatialGEV_adfun()` which returns the `adfun` object, and then `spatialGEV_fit()` which does the fitting.  The advantage is that the `adfun$env$parameters` object tells you the dimension of the parameter values, which can be useful for initialization.  Also, note that `adfun$env$data` object contains the entire data list for browsing.

- [ ] Write some tests for the `spde` kernel.  Construct the sparse precision matrix using `get_spde_prec()`, invert it using `Matrix::solve()`, apply the scale factor, and pass as variance to `dmvnorm()`.
